Tracking $p$-adic precision

### 2014-05-26 padicprec.tex
Changes for final copy, including removing most of Appendix B

\title{Tracking $p$-adic precision}
%\author{Xavier Caruso, David Roe \& Tristan Vaccon}
\date\today

@@

In Appendix \ref{sec:manifold}, we extend the results of Section \ref{sec:mainlemma}
to $p$-adic manifolds, describing how to specify precisions for points on elliptic curves
and Grassmannians.  Finally, Appendix \ref{sec:differentials} describes how to compute
the derivative of many common operations on polynomials and matrices, with an eye
toward applying Lemma \ref{lem:main}.

@@

\begin{rmk}
From the above discussion, one can easily derive a numerically stable 
algorithm that computes the SOMOS 4 sequence: 
\begin{enumerate}
\item compute the Laurent polynomials $P_n$ using the recurrence in 
the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$
\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
\end{enumerate}
However, computing the $P_n$'s is very time-consuming since it requires 
division in a polynomial ring with $4$ variables and the
size of the coefficients of $P_n$ explodes as $n$ grows.

@@

Let us first recall that a square matrix $M$
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ are lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{ssec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the Cramer-type formulae
in the second algorithm yield a substantially longer running time than the first.
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

@@

The following list gives examples of useful precision types.  A description of the $\round$ function has been omitted for brevity.
\begin{itemize}
\item The \emph{lattice} precision type has $\T = \Lat(E)$.
\item In the \emph{jagged} precision type, $\T$ consists of lattices of the shape $B_E((e_i),(r_i))$ for a fixed Banach basis $(e_i)$ of $E$.
\item In the \emph{flat} precision type, $\T$ consists of lattices $B_E(r)$.  The flat precision type is useful since it takes so little space to store and it easy to compute with.
\item If $E = K_{<d}[X]$ is the space of polynomials of degree less than $d$, the \emph{Newton} precision type consists of lattices $B_E((X^i),(r_i))$ where $-\log r_i$ is a convex function of $i$.  The Newton precision type is sensible if one thinks of polynomials as functions $K \to K$, since extra precision above the Newton polygon never increases the precision of an evaluation.
\item If $E = M_{m \times n}(K)$, the \emph{column} precision type consists of lattices with identical image under all projections $\pr_i : E \to K^m$ sending a matrix to its $i$th column.  It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.  It is important when working with overconvergent modular symbols since these lattices are stable under certain Hecke operators.
\end{itemize}

Note that sometimes the precision of a final result can be computed
a priori (using the methods of Appendix \ref{sec:differentials} for example).
Taking advantage of such knowledge can minimize artificial precision loss
even when using rougher precision types such as flat or jagged.
Separating precision from approximation also makes
it much easier to implement algorithms capable of processing different precision types,
since one can implement the arithmetic of the approximation separately from the logic
handling the precision.

@@

We now prove that it is correct.
We introduce the function $f : \Qp^\times \times \Q_p^3 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$. For all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is  
differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix
$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3)$.
Thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
$\Z_p^4$.
We are now going to prove by induction on $i$ that, at the end of the 
$i$-th iteration of the loop, we have $\prec = N + v_p(\det D_i)$ and
\begin{equation}
\label{eq:congrSOMOS}
(a, b, c, d) \equiv (u_i, u_{i+1}, u_{i+2}, u_{i+3}) \pmod
{p^N \varphi_i(\Z_p^4)}.
\end{equation}
The first point is easy. Indeed, from $D_i = D(u_{i-1}, u_i, u_{i+1}, 
u_{i+2}) D_{i-1}$, we deduce $\det D_i = \det D_{i-1} \cdot 
\frac{u_i}{u_{i-3}}$ and the assertion follows by taking determinants
and using the induction hypothesis. 
Let us now establish \eqref{eq:congrSOMOS}. To avoid confusion, 
let us agree to denote by $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
$a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
the $i$-th iteration of the loop. By induction hypothesis (or by 
initialization if $i = 1$), we have:
\begin{equation}
\label{eq:congrSOMOS2}
(a', b', c', d') \equiv (u_{i-1}, u_i, u_{i+1}, u_{i+2}) \pmod
{p^N \varphi_{i-1}(\Z_p^4)}.
\end{equation}
Moreover, we know that the determinant of $\varphi_{i-1}$ has valuation
$\prec'$. Hence \eqref{eq:congrSOMOS2} remains true if $a'$, $b'$, $c'$
and $d'$ are replaced by other values which are congruent to them modulo
$p^{\prec'}$. In particular it holds if $a'$, $b'$, $c'$ and $d'$ denotes
the values of $a$, $b$, $c$ and $d$ after the execution of line
\ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
\ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
$(u_0, u_1, u_2, u_3)$), we get:
$$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).$$
By the discussion above, this implies in particular that $f(a',b',c',d')$
belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
conclude by remarking first that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
{p^\prec \Z_p^4}$ by construction and second that $p^\prec \Z_p^4 \subset
p^N \varphi_i(\Z_p^4)$.

@@

>>>>

\title{Tracking $p$-adic precision}
\author{Xavier Caruso, David Roe and Tristan Vaccon}
\date\today

@@

In Appendix \ref{sec:manifold}, we extend the results of Section \ref{sec:mainlemma}
to $p$-adic manifolds, describing how to specify precisions for points on elliptic curves
and Grassmannians.  Finally, Appendix \ref{sec:matrices} describes how to compute
the derivative of some common operations on matrices as an example, with an eye
toward applying Lemma \ref{lem:main}.

@@

\begin{rmk}
From the above discussion, one can easily derive a numerically stable 
algorithm that computes the SOMOS 4 sequence: 
\begin{enumerate}
\item compute the polynomials $P_n$ using the recurrence in 
the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$
\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
\end{enumerate}
However, computing the $P_n$'s is very time-consuming since it requires 
division in a polynomial ring with $4$ variables and the
size of the coefficients of $P_n$ explodes as $n$ grows.

@@

Let us first recall that a square matrix $M$
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ are lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{sec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the Cramer-type formulae
in the second algorithm yield a substantially longer running time than the first.
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

@@

The following list gives examples of useful precision types.  A description of the $\round$ function has been omitted for brevity.
\begin{itemize}
\item The \emph{lattice} precision type has $\T = \Lat(E)$.
\item In the \emph{jagged} precision type, $\T$ consists of lattices of the shape $B_E((e_i),(r_i))$ for a fixed Banach basis $(e_i)$ of $E$.
\item In the \emph{flat} precision type, $\T$ consists of lattices $B_E(r)$.  The flat precision type is useful since it takes so little space to store and it easy to compute with.
\item If $E = K_{<d}[X]$ is the space of polynomials of degree less than $d$, the \emph{Newton} precision type consists of lattices $B_E((X^i),(r_i))$ where $-\log r_i$ is a convex function of $i$.  The Newton precision type is sensible if one thinks of polynomials as functions $K \to K$, since extra precision above the Newton polygon never increases the precision of an evaluation.
\item If $E = M_{m \times n}(K)$, the \emph{column} precision type consists of lattices with identical image under all projections $\pr_i : E \to K^m$ sending a matrix to its $i$th column.  It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices of the form $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.  These lattices are stable under certain Hecke operators, which is necessary for computing with overconvergent modular symbols.
\end{itemize}

Note that sometimes the precision of a final result can be computed
a priori (using the methods of Appendix \ref{sec:matrices} for example).
Taking advantage of such knowledge can minimize artificial precision loss
even when using rougher precision types such as flat or jagged.
Separating precision from approximation also makes
it much easier to implement algorithms capable of processing different precision types,
since one can implement the arithmetic of the approximation separately from the logic
handling the precision.

@@

We now prove that it is correct.
We introduce the function $f : \Qp^\times \times \Q_p^3 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$. For all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is  
differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix
$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3)$.
Thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
$\Z_p^4$.
We are now going to prove by induction on $i$ that, at the end of the 
$i$-th iteration of the loop, we have $\prec = N + v_p(\det D_i)$ and
\begin{equation}
\label{eq:congrSOMOS}
(a, b, c, d) \equiv (u_i, u_{i+1}, u_{i+2}, u_{i+3}) \pmod
{p^N \varphi_i(\Z_p^4)}.
\end{equation}
The first point is easy. Indeed, from $D_i = D(u_{i-1}, u_i, u_{i+1}, 
u_{i+2}) D_{i-1}$, we deduce $\det D_i = \det D_{i-1} \cdot 
\frac{u_i}{u_{i-3}}$ and the assertion follows by taking determinants
and using the induction hypothesis. 
Let us now establish \eqref{eq:congrSOMOS}. To avoid confusion, 
let us agree to denote by $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
$a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
the $i$-th iteration of the loop. By induction hypothesis (or by 
initialization if $i = 1$), we have:
\begin{equation}
\label{eq:congrSOMOS2}
(a', b', c', d') \equiv (u_{i-1}, u_i, u_{i+1}, u_{i+2}) \pmod
{p^N \varphi_{i-1}(\Z_p^4)}.
\end{equation}
Moreover, we know that the determinant of $\varphi_{i-1}$ has valuation
$\prec'$. Hence \eqref{eq:congrSOMOS2} remains true if $a'$, $b'$, $c'$
and $d'$ are replaced by other values which are congruent to them modulo
$p^{\prec'}$. In particular it holds if $a'$, $b'$, $c'$ and $d'$ denotes
the values of $a$, $b$, $c$ and $d$ after the execution of line
\ref{line:lift}. Applying Lemma \ref{lem:main} and Proposition
\ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
$(u_0, u_1, u_2, u_3)$), we get:
\[
f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).
\]
By the discussion above, this equation implies in particular that $f(a',b',c',d')$
belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
conclude by remarking that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
{p^\prec \Z_p^4}$ by construction and that $p^\prec \Z_p^4 \subset
p^N \varphi_i(\Z_p^4)$.

@@

\section{Example: Matrices}
\label{sec:matrices}

We saw in the core of the article that the differential of an operation 
encodes the intrinsic loss/gain of precision when performing this 
operation. In this appendix we compute the differential of various
common operations on matrices. Surprisingly, 
we observe that all differentials we will consider are rather easy to 
compute even if the underlying operation is quite involved.

\affiliationone{
Xavier Caruso and Tristan Vaccon \\
Institut de recheche en math\'ematiques de Rennes (IRMAR) \\
Universit\'e Rennes 1 \\
Campus de Beaulieu \\
35042 Rennes Cedex \\
France}
\affiliationtwo{
David Roe \\
Department of Mathematics \\
University of Calgary \\
2500 University Dr NW \\
Calgary AB T2N 1N4 \\
Canada}
